Structuring College Access: College Board Geo-markets and the Market Segment Model

### 2024-07-08 text/old_spring2024/intro.qmd
initial commit, copying files from messed up repo

In 1983, Zemsky and Oedel authored *The Structure of College Choice*, published by *The College Board.* This book develops the Market Segment Model and introduces local "community-based enrollment markets" (p.14) -- later called "geomarkets" -- which are the basis for the College Board market Enrollment Planning Service (EPS), a market research product. @RN4982[pp., 9-10] write that the Market Segment Model sought to quantify the knowledge that college admissions officers have about student demand:

@@

Scholarship in the sociology of education describes schools and colleges as "sorting machines" that reproduce inequality by incorporating, creating, and allocating students to different categories [@RN4873]. Given the consequences subsequent outcomes, the sociology of education is particularly concerned with the transition from high school to college. Over the last decade or so, scholarship on college access has begun to take more seriously the idea that colleges are not passive recipients of applications; rather they expend substantial resources soliciting demand from desirable applicants [@RN3519; @RN4774; @RN4324]. Most scholarship on recruiting analyzes the behavior of individual colleges, thereby assuming that recruiting behavior is a function of individual colleges [e.g., @RN4758; @RN4759]. @RN5012 analyze the College Board *Student Search Service* product, highlighting the role that third-party products play in structuring the recruiting behavior of colleges. The authors show that the utilization of particular search filters -- particularly geographic filters, such as zip code -- is associated with systematic racial exclusion that does not depend on which college purchased the list.

However, prior research has not examined the role of third-party market research in reproducing socioeconomic and racial inequality. This manuscript analyzes the Market Segment Model and College Board Geomarkets as a case study of quantification. We develop arguments from the broader literature on quantification [e.g.,] [@RN5003; @RN5005; @RN5007], particularly the discussions of correlation and homophily from @RN4975. Predictive analytics are based on correlation. When we use data on past correlations to make recommendations about the future, we amplify the effects of historic structural inequality. The Market Segment Model analyzes SAT score-sending data from 1980 and concludes that student demand for higher education is primarily a function of social class= [@RN4982]. Homophily is the idea that actors that share common characteristics are likely to form connections with one another. The Market Segment Model argues that homophily is the organizing principle of student demand and competition between colleges, that "the hierarchical structure of collegiate competition largely reflects the stratified social and economic dimensions of the communities from which colleges draw their students" [@RN4982, p. 72]. Subsequently, the assumption of homophily is programmed into the EPS and Student Service Service products that colleges utilize to decide where to focus recruiting energy.

Empirically, our analyses address the following research questions, which speak to how geomarkets are utilized within the Enrollment Planning Service and within the Student Search Service student list product. First, what is the socioeconomic and racial variation between geomarkets in metropolitan areas and how does this variation change over time? We address this question by spatially joining the geomarket shapefile to Census data about socioeconomic and racial characteristics. Second, how does the socioeconomic and racial composition of included versus included prospects vary when student list purchases filter on particular geomarkets, in concert with other commonly used search filters? We address this question by analyzing student list purchases that do not filter geomarkets and simulating which prospects would have been included had they filtered on particular geomarkets. We obtained these data by issuing public records requests to public universities in four states. Third, in student list purchases that filtered on geomarkets, how does the socioeconomic and racial composition of included students compare the that of the geomarket and how does it compare to that of the broader metropolitan area?

@@

# Background: Enrollment Management

The term "enrollment management" (EM) refers to a profession, an administrative structure, and an industry. As a profession, EM integrates techniques from marketing and economics in order to "influence the characteristics and the size of enrolled student bodies" [@RN2771, p. xiv]. As an administrative structure, the EM office typically controls the activities of admissions, financial aid, and recruiting [@RN2406]. The EM industry consists of university personnel (e.g., admissions counselors, VP for enrollment management), professional associations (e.g., National Association for College Admission Counseling), and third-party servicers that produce market research, software, and and consulting services (e.g., College Board, ACT, EAB, Huron, Slate, PowerSchool).

Figure @fig-em-funnel depicts the "enrollment funnel," which modifies the traditional "marketing funnel" to depict broad stages in the process of recruiting students [@RN4986; @RN2004]. The funnel begins with a large pool of "prospects" (i.e., prospective students) that the university would like to enroll. "Leads" are prospects whose contact information has been obtained. "Inquiries" are prospects that contact the institution and consist of two types: first, inquiries who respond to an initial solicitation (e.g., email) from the university; and second, "student-as-first-contact" inquiries who reach out to the university on their own (e.g., sending ACT scores). <!--  Applicants consist of inquiries who apply plus "stealth applicants" who do not contact the university before applying.  -->The funnel narrows at each successive stage in order to convey the assumption of "melt" (e.g., a subset of "inquiries" will apply). Practically, the purpose of the enrollment funnel is to inform recruiting interventions that target one or more stages. These interventions seek to increase the probability of "conversion" across stages [@RN4322]. At the top of the enrollment funnel, purchasing student lists is the primary means of converting prospects to leads [@list_biz]. Purchased leads are served emails, brochures, and targeted social media advertisements designed to solicit inquiries and applications [@RN4895].

Scholarship at the nexus of enrollment management and college access can be categorized by which part(s) of the enrollment funnel they focus on. The majority of scholarship focuses on the admissions stage, analyzing which admissions criteria are utilized and/or which applicants are admitted [e.g., @RN4874; @RN4993; @RN4131; @RN3522; @poon2022rethinking]. A growing literature analyzes the earlier "recruiting" stages of identifying leads, soliciting inquiries, and soliciting applications. Scholarship on recruiting has conceptualized recruiting behavior as an indicator of college enrollment priorities [@RN4758] and identifies recruiting practices as a mechanism for social reproduction in college access [@RN4324; @RN3519].

Ethnographies by @RN3519 and by @RN4407 analyze the connections between privileged colleges and privileged high schools. @RN4407 shows how guidance counselors at elite private school guidance counselors lobby admissions counselors on behalf of marginal students. @RN3519 finds that college admissions officers visit high schools as a means of maintaining positive relationships with guidance counselors at affluent feeder schools. Consistent with this finding, quantitative case studies of off-campus recruiting visits show that selective private colleges visit affluent, predominantly white schools and communities, with a disproportionate number of visits to private high schools [@RN5014; @RN4330]. Additionally, public research universities often make make more out-of-state recruiting visits than in-state visits, and these out-of-state visits focus on the same set sorts of schools targeted by selective private universities [@RN4759; @RN4758].

Other studies investigate recruiting behavior by non-selective institutions [@RN4520; @cottom2017lower; @dache2018dangling]. Analyzing recruiting from the perspective of high school students, @RN4324 finds that underrepresented students were drawn to colleges that made them feel wanted, often attending institutions with lower graduation rates and requiring larger loans than other college options. @cottom2017lower shows that for-profit colleges found a niche in Black and Latinx communities because traditional colleges ignored these communities. <!--  WHAT YOU SAID ABOUT POSECZNICK Posecznick (2017) studied the “Ravenwood College,” a private nonprofit college founded to serve working Black women. Consistent with analyses of tuition discounting by Lapovsky (2001), the survival of Ravenwood depends on the success of admissions officers in identifying prospects (e.g., through on- and off-campus recruiting events) and “converting” these prospects to enrolled students.  -->

We make two observations about the recruiting literature. First, consistent with scholarship on political economy [@RN4517; @cottom2017lower; @RN5015] and industrial organization [@RN1549; @RN2133], the industry is characterized by a vertical hierarchy of postsecondary institutions matching to a vertical hierarchy of customer niches, a process that exemplifies the idea that higher education is mechanism of social reproduction [@RN2377; @RN4873].<!--  For example, for-profit institutions target adults in communities not being targeted by traditional colleges [@cottom2017lower]. Non-selective private non-profits target prospects who might be considering community college or a local state colleges [@RN4324]. Selective private colleges target private high schools, with a particular link between sectarian high schools and colleges of the same faith [@RN5014], but these schools are increasingly targeted by out-of-state public research universities, suggesting growing competition for a coveted market niche. --> Second, by focusing on the behaviors of individual colleges, the literature implicitly states that recruiting is something done by individual colleges. 
<!--  
Why then, does recruiting behavior look so similar across colleges of a particular "type"?[^1]
[^1]: New institutional theory offers one explanation [@RN527]. In highly developed organizational fields like higher education, forces of competitive and institutional isomorphism -- including institutional locics -- compel organizations adopt similar practices as their peers. Practices deemed advantageous spread quickly, for example, public research universities targeting nonresident prospects to compensate for declines in state appropriations. These behaviors are conceived as autonomous decisions by individual colleges, albeit structured by institutional logics.
 -->

<!-- make or buy  
The process of recruiting students involves many "make or buy" decisions about whether to perform a given task in-house or outsource it to a third-party vendor. EM consulting firms provide advice and implementation in the areas of marketing, recruiting, pricing and financial aid, and student success. As "creating a class" has becomes complicated and high-stakes [@RN3519], many colleges hire EM consulting firms to develop and/or implement recruiting campaigns. The two largest enrollment consulting firms -- Ruffalo Noel Levitz and EAB -- claim to serve more than 3,000 colleges and universities collectively [@rnlabout; @eabcomm]. In our data collection, attempting to obtain data about student list purchases from all public universities in four states, at least half of these universities outsourced student list purchases to an EM consultancy [@list_biz; @list_empirics].
-->

However, the recruiting behavior of colleges is structured by third-party actors. Drawing from scholarship on organizational theory [@RN4738], enrollment management processes involve many "make or buy" decisions about whether to perform a given task in-house or outsource it to a third-party vendor [@list_biz]. EM consulting firms provide advice and implementation in the areas of marketing, recruiting, pricing and financial aid, and student success. <!-- As "creating a class" has becomes complicated and high-stakes [@RN3519], many colleges hire EM consulting firms to develop and/or implement recruiting campaigns. --> The two largest enrollment consulting firms -- Ruffalo Noel Levitz and EAB -- claim to serve more than 3,000 colleges and universities collectively [@rnlabout; @eabcomm].

Moreover, process of college access is structured by third-party recruiting products utilized by colleges. @RN5012 deconstruct the College Board *Student Search Service* product. Selection devices are procedures or routines for making categorical decisions based on input values [@RN4778]. Student list products are selection devices that enable colleges to choose which prospective students they target or ignore based on search filters that incorporate search filters (e.g., high school graduating class, state) which enable colleges to select prospective students. "Racialized inputs" are ostensibly race-neutral inputs that are systematically correlated with race because marginalized racial/ethnic groups have historically been excluded from the input [@RN4786]. @RN5012 argue that several frequently utilized student list filters (zip code, AP test score, SAT score) are racialized inputs. Using a national sample of high school students, and also using data from actual student lists purchased by public universities, they show that racialized search filters have a strong negative relationship with the selection of Black and Latinx prospects. 
 <!-- Using data from actual student lists purchased by public universities, they show that raciaized search filters -- used in conjunction with other search filters -- yield dramatic racial inequality in which prospective students are targeted. -->

Student list products offer many geographic filters. Geographic borders are perhaps the most commonly studied racialized input in scholarship about algorithmic bias [@RN4775; @RN4849; @RN4887], owing to extreme racial segregation in the U.S. In addition to offering geographic filters based on known geographic borders (e.g., zip code, county, CBSA, state), the College Board created new geographic borders that subsequently became filters in the *Student Search Service*. The geodemographic segment filters utilize cluster analysis to allocate each census tract and each high school into different categories based past college enrollment and other factors [@RN4565]. Geodemographic search filters can be read as the mimicry of corporate market logic [@RN4004] in that they apply the logic and methods of commercial geodemographic market segmentation products, such as PRIZM by the Claritas Corporation [@RN4957], to the case of college recruiting.

Geomarket search filters, the focus of this manuscript, slice states and metropolitan areas into smaller, local recruiting markets. For example, geomarket "CA08" refers to "Alameda Country excluding Oakland." A university might purchase prospect contact information by filtering on this geographic region in combination with additional filters, such as year of high school graduation and SAT test score. When we initiated this manuscript, we asked ourselves, "where did geomarket filters come from?" The answer surprised us. We had imagined ourselves studying the application of the "data science revolution" to college recruiting, but it turned out we were studying the 1970s and 1980s! The Market Segment Model and geomarkets can be read as the application of social science research methods to "capture and quantify the phenomenon behind the folklore" [@RN4982, p. 11] of college admissions officers. This work, comissioned by College Board, was subsequently turned into a market research tool -- The College Board Enrollment Planning Service -- that helps colleges decide where to look for prospective students.

```{=html}
<!--The enrollment management industry emphasizes market research.
Market research is the process of producing information that informs decisions about product design or how to find customers.   -->
```

The enrollment management industry puts great stock in market research. @RN4986[p. 17] define market research as all "activities intended to increase understanding of a market and consumer behavior in that market." At the most basic level, in-house market research is a function of being intentional about record-keeping and using those records [@RN3519].^[@RN3519 describes the archives of annual reports of admissions activities from a selective liberal arts college. In the 1942 report, "Section 2.5 is titled 'Secondary School Quartile Standings," a measure of where matriculants fell in the rankings of their high school classes" [@RN3519] while Section 3.11 describes the high schools visited. In the chapter entitled "Travel," @RN3519 [p. 69] describes how The College used record-keeping to inform decisions about visits to local high schools in the 2000s: "The College keeps files on hundreds of schools across the country, and around the globe, with which it had contact \ldots The files enable officers to quickly discern"how we've done here lately," \ldots with the number, relative quality, and yield rate of applicants from a particular school."] However, most market research in the contemporary enrollment management industry is produced by third-party organizations. For example, Ruffalo-Noel Levitz and EAB both produce prodigious amounts of market research, some publicly available as a means of business development [e.g., @RN4741; @RN4728, but most only available to clients. EPS remains the market research arm of College Board, while Encoura Eduventures is the market research arm of ACT.

Despite being central to enrollment management for at least 40 years [@RN4986], the role of third-party market research has been ignored by scholarship on college access. Scholarship on college access often takes a demand-side perspective, whereby household differences in economic and cultural capital yield stratification in student preferences and the ability to gain admission [e.g., @RN4935;@RN1814]. However, market research informs where colleges look for students and which prospective students to target. These recruiting interventions intervene on student demand [@RN4324; @cottom2017lower], thereby structuring college access pathways in ways that could amplify or diminish historic inequality in college access. The College Board played a pivotal role in transforming market research about enrollment from something done by individual universities to something done by third-party professionals [@RN4986]. This article offers one story about the quiet revolution that transformed college access into a sophisticated, supply-led enterprise. We argue that market research structures who colleges recruit, in ways that reinforce existing social stratification. The case of Market Segment Model and College Board geomarkets shows how market research -- and its commodification into recruiting products -- is a mechanism for social reproduction. 
[LAST TWO SENTENCES ARE REDUNDANT BUT BUT ARE TRYING TO DO SLIGHTLY DIFFERENT THINGS AND I LIKE THEM BOTH]

# Quantification

```{=html}
<!--
CUT GENERAL PARAGRAPH ON REPRODUCTION

Within the sociology of education, @RN2377 describes education as a sieve, allocating people to opportunities in socially stratified, capitalist societies. This role highlights "Max Weber's core insight that education has a dual character" [@RN2377, p. 129], on one hand, serving social mobility by redistributing access to privileged positions and, on the other, serving social reproduction by granting privileged families disproportionate access to the educational institutions that confer pathways to privileged positions. Similarly, @RN4873 describe schools and colleges as "sorting machines" that incorporate external categories (e.g., race, class), create categories (e.g., instructional tracks), and allocate students to categories. Initial category allocation affects subsequent allocation, such that school assignments and instructional track assignments affect college access, which affects employment assignments. "In evaluating schools as inequality regimes" @RN4873 \[p. 322\] argue, "the relevant question is not whether schools produce inequality; rather, the relevant questions are, when and where do schools produce inequality, \[and\] how do they do so." As @RN2377 show, whereas early research on social stratification sought to identify patterns of inequality [e.g., @RN4873; @RN290; @RN1268], the sociology of education is now more concerned with identifying the mechanisms underlying these patterns. 
-->
```
We situate our study amidst scholarship on within scholarship on quantification as a mechanism that reproduces inequality. @RN5007 [p. 402] define quantification as "the production and communication of numbers" (p. 402) and introduce reacivity, discipline, and authority as three interrelated themes that describe the effects of quantification. Reactivity is the idea that salient quantitative measures cause people and organizations to change their behavior. Discipline is the idea that quantification causes actors to react in a particular way. Authority is the idea that quantification changes decision-making power, often weakening the discretion of local decision-makers.

Scholarship about U.S. News & World Report (USNWR) Law School rankings demonstrate the reactivity, discipline, and authority effects of quantification [@RN5003; @RN2356; @RN2376; @RN2375]. Developed as a means of informing prospective students about quality, rankings had the intended effect of affecting. Rankings also affected hiring decisions of firms because clients evaluated firms based on the prestige of law schools attended by firm lawyers. Once law schools realized that "important groups of constituents --- students, faculty, trustees, employers, other media --- were using rankings to make decisions that had large consequences for schools \ldots \[then\] schools felt pressured to take them seriously" [@RN5007, p. 415]. Rankings *disciplined* the behavior of law schools. For example, applicants with characteristics valued by the rankings system (e.g., LSAT scores) became more important for decisions about admissions and merit aid [@RN5003]. Rankings also weakened the *authority* of admissions directors and admissions officers to make independent decisions about which applicants to admit, ideal class size, and curricular offerings [@RN5003]. <!--
In a process akin to sports analytics in Moneyball \[CITE MICHAEL LEWIS\], we suggest that market research replaces the wisdom and authority of local admissions counselors.    -->

The analysis of UK League Tables by @RN5005 show how quantification can be a mechanism of social reproduction. As part of a broader *New Public Management* effort to infuse market competition within public providers [@RN161], the UK government began publishing school "league tables" in 1992. League tables ranked schools based on student performance on high-stakes national exams taken at age 16. Using Census data on occupational class measured at the local authority level (similar to a U.S. county), @RN5005 find that localities with higher performing schools experienced growth in the share of managerial/professional households following the adoption of league tables and a decline in the share of working-class households. Additionally, using longitudinal household survey data, managerial/professional households were more likely to move to localities with higher ranked schools following the introduction of school league tables. With respect to mechanisms, league tables facilitate making evaluative comparisons between schools regardless of geographic proximity. They also discipline households and schools to conceive of school quality in terms of test scores, which are substantially a function of the class composition of schools. Compared to working-class parents, parents with professional/managerial occupations are more aware of school league tables and have financial resources necessary to move to neighborhoods near a "good" school. Therefore, league tables contributed to social reproduction because managerial/professional households respond to consumer-facing metrics by moving to more expensive neighborhoods, near higher performing schools. <!-- THIS IS RATCHET EFFECT STUFF. I CUT IT BECAUSE YOU HAVEN'T INTRODUCED THE IDEA OF RATCHET EFFECT
Schools that performed well on initial league tables tended to be schools with a high share of professional/managerial families. Therefore, quantification takes a snapshot of existing social stratification and that snapshot becomes a power unto itself, amplifying the effect of previous strctural inequality on subsequent structural inequality.
-->

```{=html}
<!-- THIS IS THE CUT TRANSITION PARAGRAPH
In the broader quantification literature, critical data studies [e.g., @RN4975; @RN4849; @RN4772] and sociological analyses of technology and inequality [e.g., @RN4778; @RN4786; @RN4877; @RN4810] introduce concepts salient to the analysis of market research. Our focus is on "selection devices," which utilize some combination of inputs and discretion to categorize places, people, and organizations for the purpose of allocating opportunities or interventions [@RN4778; @RN5012].
-->
```
In the interdisciplinary literature on quantification [@RN5008], the discussions of correlation and homophily by @RN4975 introduce concepts that are salient to the analysis of market research. Chapter 1 of @RN4975 is *Correlating Eugenics*. Correlation measures the extent to which two or more variables move together. The Pearson Correlation Coefficient was developed by the 19th Century statisticians and "biometric eugenicists" Francis Galton and his student Karl Pearson [@RN4975]. @RN4975 [p. 50] writes that "correlation grounds big data's so-called revolutionary potential. As Wired editor Chris Anderson infamously declared in his 2008 editorial 'The End of Theory,' big data proved that 'correlation supersedes causation, and science can advance even without coherent models, unified theories, or really any mechanistic explanation at all.'" Predictive analytics are developed in two steps. First, apply statistical techniques to previous cases (training data) in order to identify factors positively and negatively associated with an outcome of interest. Second, apply these results (e.g., regression coefficients) to future cases in order to make predictions and to assign levels of risk to each case. These models predict outcomes based on correlations without requiring knowledge about underlying causal relationships. @RN4975 provides the example of @RN4989, who develop a method to predict sensitive personal attributes (e.g., gender, political party) based on Facebook Likes.

Due to data limitations, predictive analytics often utilize "proxy" variables [@RN4849; @RN4975], which are variables that are correlated to a variable of interest. For example, "e-scores" utilize proxy variables to identify "people like you" and then predict your buying behavior based on the past buying behavior of these others. @RN4849 [p. 146] states that "the modelers for e-scores have to make do with trying to answer the question 'How have people like you behaved in the past?' when ideally they would ask, 'How have you behaved in the past?'" We observe similar behavior in market research about college access, when enrollment managers use 'which colleges did students near you consider' as a proxy for, 'which colleges would you consider?'

Many studies show that using correlational models make predictions reproduces structural inequality. The correlations observed during the training data stage are a snapshot of relationships between variables at a particular point of time. These correlations may be a function of enduring structural inequality, but underlying causes are not considered by applications of predictive models. @RN4794 [p. 224] state that "predicting the future on the basis of the past threatens to reify and reproduce existing inequalities." Therefore, disproportionately targeted/excluded populations are predicted to have a higher risk of an outcome, which amplifies disproportionate targeting/exclusion. This phenomenon whereby has been termed the "ratchet effect" [@RN4851] and "pernicious feedback loops" [@RN4849]. An often cited example is the LSI-R recidivism model which predicts a prisoner's chances of re-arrest and is used by 24 states [@RN4849]. Because the algorithm uses zip code as an input, people who live in highly policed neighborhoods have a higher predicted probability of being arrested, which leads to more policing in those neighborhoods, which perpetuates racialized inequality in arrests. Note that predictive models such as the LSI-R model are not merely used for social science purposes. Rather, they reproduce structural inequality because they direct the allocation of future resources based on a snapshot of historical inequality.

Chapter 2 of @RN4975, *Homophily, or the Swarming of the Segregated Neighborhood*, argues that homophily often underlies the use of correlation to predict future behavior. Homophily means that actors who share common characteristics are likely to form connections with one another. The ubiquitous colloquialism is "birds of a feather flock together." Homophily is a core assumption for network science [@mcpherson2001birds], in which actors (e.g., high school students) are linked to one another through direct and indirect network ties (e.g., submit SAT scores to the same college). Because network science often draws from rational choice theory, it assumes that homophily is the result of voluntary action by individuals, thereby "eras\[ing\] historical contingencies, institutional discrimination, and economic realities" [@RN4975, p. 95] that underlie behavior consistent with homophily. @RN4975 problematizes the idea that homophily is a naturally occurring phenomenon. In commercial social networks, such as Facebook, Twitter/X, and TikTok, homophily is programmed into the algorithms that create connections between users.[^4]

[^4]: Thus, @RN4975 [p. 82] writes, "echo chambers are not unfortunate errors, but deliberate goals" because "homophily is used to create agitated clusters of individuals whose angry similarity and overwhelming attrication to their common object of hatred both repel them from one another and glue them together."

Within market research, homophily is central to geodemography and the creation of market segments. Geodemography emerged in the 1970s as a branch of market research that estimates the behavior of consumers based on where they live [@RN4878]. Market segments are subgroups within a larger market that have similar consumer demand. Early geodemographic classifications of consumers (e.g., PRIZM by Claritas Corporation) were derived from publicly available Census data, which disaggregated data to the zip code level [@RN4957]. The Claritas Corporation argued -- and had a financial incentive to do so -- that people living near one another share similar consumer preferences. Therefore, geographic localities could be categorized into market segments that would be useful for direct mail marketing campaigns. Later, the development of individual credit scores (e.g., FICO score) enabled merchants to classify consumers into many, fine-grained groups [@RN4877]. @RN4810 introduce the concept "classification situations" to describe the expansion of actuarial techniques to categorize customers into many, ordinally ranked groups. Merchants and lenders began tying these classifications to tiered products that targets different consumer groups with different levels of benefits and costs [@RN5015]. For example, "payday loans" charge high interest rates to consumer groups that were previously denied credit altogether. Classification situations engender markets where a vertical hierarchy of products are targeted to a vertical hierarchy of consumers. <!--
ORIGINAL TEXT ABOUT GEODEMOGRAPHIC CLASSIFICATION SYSTEMS

Clusters were given evocative names like "Blue Blood Estates," "\[Archie\] Bunker's Neighbors", and "Share Croppers." Because U.S. publicly available U.S. Census data disaggregated data only to the zip code-level, the Claritas Corporation had a strong incentive to argue that consumer behavior within zip codes was very similar [@RN4957]. @RN4565 emulated the behavior of Claritas by developing the *Segment Analysis Service: An Educationally Relevant Geodemographic Tagging Service*, which was an "add-on" to the *Student Search Service* student list product. Describing its conceptual underpinnings, @RN4565 [p. 1] states, "the basic tenet of geodemography is that people with similar cultural backgrounds, means, and perspectives naturally gravitate toward one another or form relatively homogeneous communities; in other words, birds of a feather flock together." The Segment product allocated each high school and each census tract into a distinct "educational neighborhoods" thought to reflect useful market segments for colleges looking for students.

PREVIOUS TEXT ABOUT CLASSIFICATION SITUATIONS
Related to the creation of market segments, @RN4810 introduce the concept "classification situations" to describe the expansion of actuarial techniques to categorize customers into many, ordinally ranked groups. Historically, classifications were binary. Consumers with "good" credit were offered loans and those with bad credit were not. The innovation of individual FICO scores -- and, later, other credit scores -- enabled merchants to classify consumers into many, fine-grained groups or along a continuum. Merchants and lenders began tying these classifications to tiered products that targets different consumer groups with different levels of benefits and costs [@RN5015]. People previously denied credit were now granted credit on different terms. For example, "payday loans" charge high interest rates to consumer groups that were previously denied credit altogether. Thus, at the bottom of the continuum, contemporary classification situations produce "predatory inclusion" -- exemplified by payday loans and for-profit colleges -- which is the logic of "including marginalized consumer-citizens into ostensibly democratizing mobility schemes on extractive terms" [@RN4774, p. 443].

Negative consequences of homophily
Chun (2021) discusses the negative consequences arising from homophily.Second, when we “engineer homophily” – through algorithms or policies – we make it less likely for heterophily to emerge, eliminating possibilities – of friendships, of matches – that may have occurred in the absence of such engineering.
-->

Reviewing recent scholarship on quantification, @RN5006 argue that quantification has effects to the extent that stakeholders care about the numbers. Law school rankings [e.g., @RN5003] and school league tables [e.g., @RN5005] seem to exemplify this sort of salience. Widely known studies of quantification in education tend to study consumer-facing quantification. To our knowledge, prior research on quantification in education has not examined how third-party market research classifies consumers. This manuscript analyzes "market segment model" [@RN4982], which categorized high school students into vertical market segments and simultaneously created local "geo-markets" that were evaluated based on their composition of student market segments. The market segment model became the basis for the College Board *Enrollment Planning Service* (EPS), which advised colleges which geo-markets to target. @RN5021 reports that in 1995, 37% of 4-year public universities and 49% of 4-year private universities used EPS, while 41% of 4-year publics and 16% of 4-year privates used ACT's market analysis service product, which was similar to EPS. Geo-markets later became a filter in the College Board *Student Search Service* student list product, which is utilized by the vast majority of public and private universities. <!--
CUT COMMENTS ON SALIENCE
By contrast, many quantification projects -- for example, human rights indicators -- have "little oomph" [@RN5006, p. 265]. FROM BERMAN & HIRSCHAN PAGE 263: "Merry demonstrates how human trafficking indicators largely sit on the shelf because no audience for them has been created." 
--> Scholarship on predictive analytics [@RN4975; @RN4849] suggests the mechanisms by which market research reproduces structural inequality. Based on a snapshot of existing social stratification, market research matches verticially categorized consumers to vertically categorized producers, thereby amplifying the effect of initial stratification on subsequent stratification. Unlike studies like @RN5005 we cannot show the effect of quantification. However, this study provides inisight about the mechanism underlying the effect. By analyzing and simulating *Student Search Service* purchases that filter on geo-markets we can show *how* the geo-markets reproduce historical class-based and race-based inequality in college access.

# The Market Segment Model and College Board Geomarkets

 <!--Historically, College Board partnered with Members to develop projects that later became revenue-generating products for College Board.-->
As a member organization, several College Board products developed from joint collaborations with administrators or professors from member institutions. During the 1970s, under the leadership of Larry Litten, then director of institutional research, Carleton College pionered applying market research to student recruiting. In the late 1970s, College Board provided Carleton College with funding and data for a six-year project that analyzed college choice decisions by households in six metropolitan areas. This project was the basis for the landmark book *Applying Market Research in College Admissions* [@RN4986], published by College Board, and was the basis for the College Board Admitted Student Questionnaire (ASQ), a revenue-generating product that provides colleges with information about admitted students -- those who enrolled and who did not enroll. Litten would later become Associate Director of the Consortium of Financing Higher Education (COFHE), a consortium of 30 selective private universities founded in the mid-1970s.

In 1978, Robert Zemsky, a University of Pennsylvania professor, was asked by the President to figure out, "'Who thinks about Penn?'" and "'What other institutions do they think about when they think about us?" [@RN4982, p. x]. To answer these questions, Zemsky began working with the Market Research Committee of the Consortium on Financing Higher Education (COFHE). As the project project became more ambitious, the researchers reached out to College Board because "we needed a database that described most institutions and most students\ldots Coincidentally, the Board was reviewing its own efforts to help colleges estimate their enrollment potential, efforts which had faltered largely because the smallest demographic unit used in tehse analyses was the state' [@RN4982, p. x]. In 1979, College Board began providing data and funding for what became the *Comprehensive Undergraduate Enrollment Planning Project (CUEPP)*.^[say that Pennsylvania Department of Education was also a partner; see page x of Zemsky] @RN4982[p. 4] write that,

> For our analysis, we sought not a complex mathematical model, but a straightforward classification system that would track the pattern of SAT-score submissions to create a map of student choice. The Market Segment model that we developed was nothing more than a set of simple rules for disaggregating high school seniors into similar groups. The model worked because students, once so disaggregated, appeared to behave in remarkably consistent ways.

**Creating the Market Segment Model.** Considering the @RN5007 \[p. 402\] definition of quantification as "the production and communication of numbers," the Market Segment Model is an effort to quantify and communicate the knowledge that admissions officers have about student demand. @RN4982[p. 4] state that "our initial task was to define enrollment markets in a manner consistent with admissions officers' intuitive understanding of student pools." They created  "three types of boundaries" [@RN4982, p. 11], region, state, and community. The three regions were New England, Middle States, and the South. Next, "we divided each state into as few as two and as many as thirty community-based enrollment markets or pools, for a total of 143 separate markets" \[in New England, Middle States, and the South\].

These enrollment markets, later called "geomarkets" by the College Board Enrollment Planning Service were intended to be consistent with the conception of a catchment market from the perspective of admissions counselors. We know little about how geomarkets were created. @RN4982 write that,

> In many cases, the market boundaries match formal political and educational divisions, reflecting natural channels of communication. Each major metropolitan area is composed of several markets, usually corresponding to the inner city, a first ring of suburbs, and an outer ring of suburbs. In more sparsely populated areas, communities are sometimes combined in order to make the analysis meaningful."

>>>>

```{=html}
<!-- CUT ORIGINS OF MARKET SEGMENT MODEL; TAKES TOO LONG TO GET TO THE NICE QUOTES, WHICH IS SUPPOSED TO BE THE VIGNETTE PART
@RN4982 describes the origins of the Market Segment Model. In 1978, Robert Zemsky, a University of Pennsylvania professor, was asked by the President to figure out, "'Who thinks about Penn?'" and "'What other institutions do they think about when they think about us?" (p. x) To answer these questions, Zemsky began working with the Market Research Committee of the Consortium on Financing Higher Education (COFHE), an association of thirty-ish selective private universities. As the project project became more ambitious, the researchers reached out to College Board because "we needed a database that described most institutions and most students\ldots Coincientally, the Board was reviewing its own efforts to help colleges estimate their enrollment potential, efforts which had faltered largely because the smallest demographic unit used in tehse analyses was the state' [@RN4982, p. x]. In 1979, College Board began providing data and funding for what became the Comprehensive Undergraduate Enrollment Planning Project (CUEPP).^[say that Pennsylvania Department of Education was also a partner; see page x of Zemsky] [MAYBE CUT SOME TEXT OF THIS PARAGRAPH]
-->
```
In 1983, Zemsky and Oedel authored *The Structure of College Choice*, published by *The College Board.* This book develops the Market Segment Model and introduces local "community-based enrollment markets" (p.14) -- later called "geomarkets" -- which are the basis for the College Board Enrollment Planning Service (EPS), which provides market research to colleges about where to look for students. Whereas subsequent College Board products (e.g., Predictor Plus) can be read as the mimicry of corporate market logic, the Market Segment Model attempts to quantify the knowledge that college admissions officers have about student demand [@RN4982, pp. 9-10]:

@@

[TRANSITION TO SCHOLARSHIP] @RN4873 describes schools and colleges as "sorting machines" that reproduce inequality by incorporating, creating, and allocating students to different categories. Given the consequences subsequent outcomes, the sociology of education is particularly concerned with the transition from high school to college. Over the last decade or so, scholarship on college access has begun to take more seriously the idea that colleges are not passive recipients of applications; rather they expend substantial resources soliciting demand from desirable applicants [@RN3519; @RN4774; @RN4324]. Most scholarship on recruiting analyzes the behavior of individual colleges, thereby assuming that recruiting behavior is a function of individual colleges [e.g., @RN4758; @RN4759]. @RN5012 analyze the College Board *Student Search Service* product, highlighting the role that third-party products play in structuring the recruiting behavior of colleges. The authors show that the utilization of particular search filters -- particularly geographic filters, such as zip code -- is associated with systematic racial exclusion that does not depend on which college purchased the list. However, prior research has not examined the role of third-party market research in reproducing socioeconomic and racial inequality

This manuscript analyzes the Market Segment Model and College Board Geomarkets as a case study of quantification. We develop arguments from the broader literature on how quantification reproduces inequality \[e.g.,\] [@RN5003; @RN5005]. In particular, we draw from the discussions of correlation and homophily by @RN4975. Predictive analytics are based on correlation. However, when we use data from past cases to make recommendations about future cases, we amplify the effects of historic structural inequality. The Market Segment Model analyzes SAT score-sending data from 1980 and concludes that student demand for higher education is primarily a function of social class= [@RN4982]. Homophily is the idea that actors that share common characteristics are likely to form connections with one another. The Market Segment Model argues that homophily is the organizing principle of student demand and competition between colleges, that "the hierarchical structure of collegiate competition largely reflects the stratified social and economic dimensions of the communities from which colleges draw their students" [@RN4982, p. 72]. Subsequently, the assumption of homophily is programmed into the EPS and Student Service Service products that colleges utilize to decide where to focus recruiting energy.

Empirically, our analyses address the following research questions, which speak to how geomarkets are utilized within the Enrollment Planning Service and within the Student Search Service student list product. First, what is the socioeconomic and racial variation between geomarkets in metropolitan areas and how does this variation change over time? We address this question by spatially joining the geomarket shapefile to Census data about socioeconomic and racial characteristics. Second, how does the socioeconomic and racial composition of included versus included prospects vary when student list purchases filter on particular geomarkets, in concert with other commonly used search filters. We address this question by analyzing student list purchases that do not filter geomarkets and simulating which prospects would have been included had they filtered on particular geomarkets. We obtained these data by issuing public records requests to public universities in four states. Third, in student list purchases that filtered on geomarkets, how does the socioeconomic and racial composition of included students compare the that of the geomarket and how does it compare to that of the broader metropolitan area?

@@
